---
title: "Practical 4 Answers"
format: html
echo: false
---

# General Linear Models (GLMs): Working with categorical variables.

-Fit and interpret one-way ANOVA and ANCOVA models in R.\\
-Construct and apply contrast coding systems.\\
-Compare and interpret additive and interaction GLM models with factors.

```{r}
library(dplyr) 
library(readr) # read_csv().
library(ggplot2)

crickets <- read_csv("data/crickets.csv")
crickets
```

# Question 1
Recreate the plot above using a colour scale for the species rather than plotting on separate facets.
```{r}
crickets |> 
  ggplot(aes(x = temp, y = rate, color = species)) +
  geom_point() +
  scale_x_continuous("Temperature (°C)") +
  scale_y_continuous("Chirp rate") +
  scale_colour_discrete("Species") +
  ggtitle("Cricket chirp rate vs temperature by species") +
  theme_bw()
```
**Colouring by species (rather than faceting) makes it much easier to see how the two species compare**

# Question 2
Fit the one-way ANOVA model rate ~ species.
```{r}
  anova_mod <- lm(rate~species, data = crickets) 
  summary(anova_mod)
```
**Note: A one-way ANOVA is a linear regression where the predictor is categorical.**

# Question 3
Which level of species has been chosen as the reference category? How can you tell from the lm() output?

O.exclamationis is the reference category. **In R, lm() defaults to treatment (dummy) coding for factors, which**
**makes the first level of the factors as the baseline. Since it is dummy coding**
$\beta_0$ or (Intercept) is the mean estimate for the reference group (O.exclamationis). 
$\beta _1$ or (speciesO.niveus) is the mean difference between speciesO.niveus and the reference group.

# Question 4
Write out the full fitted model equation, defining all terms clearly.

$$\mathrm{rate}_i= 85.59 - 23.16 \times [\mathrm{species}_i = \mathrm{O.niveus}]$$

$\mathrm{rate}_i$ is the predicted chirp rate for observation $i$.
$1[\cdot]$ is an indicator function that is $1$ if cricket $i$ is O.$\mathrm{niveus}$ and 0 if not.(O.$\mathrm{exlamationis}$) is the reference level.

The mean rate for O.$\mathrm{exlamationis}$ is 85.59 and the mean rate for O.$\mathrm{niveus}$ is 23.16 lower, which is
$85.59 - 23.16 = 62.43$.

# Question 5
Calculate the estimated mean rate for each species using only the summary of the model. Confirm your results with the predict() function.

```{r}
# using coefficients
intercept <- coefficients(anova_mod)["(Intercept)"]
niveus_offset <- coefficients(anova_mod)["speciesO. niveus"]

exclamationis_mean <- intercept
niveus_mean <- intercept + niveus_offset # intercept - 23.16

c(exclaimationis = exclamationis_mean, niveus = niveus_mean) # displays the reference group and estimated mean
```
```{r}
# Confirm with predict()
predict(anova_mod, newdata = tibble(species = c("O. exclamationis", "O. niveus")))
```
**Both approaches agree: an estimated mean chirp rate of 85.59 for O. exclamationis and 62.43 for O. niveus. These are each species’ raw sample mean, since a one-way ANOVA with a single categorical predictor and no other covariates estimates group means directly.**

# Question 6
Refit the model without a reference category. What sort of coding is this? Are the results the same?

```{r}
anova_mod_nointercept <- lm(rate~species-1,data = crickets)
summary(anova_mod_nointercept)
```
**Removing the intercept (by including - 1) switches from treatment/dummy coding to means/cell-means coding.**

Comparing the two: **treatment/dummy coding**: intercept (reference) plus an offset(s).
**means coding**: each level of species gets its own coefficient, which is that group’s mean.

Note that the underlying model (including fitted values, residuals, $R^2$) is identical. The only change is the parameterisation.**

# Question 7
Add the temperature term to the regression. Write out the mathematical equation for this model. Is this an additive ANCOVA model?

```{r}
  ancova_mod <- lm(rate ~ species + temp, data = crickets)
  summary(ancova_mod)
```
$$\mathrm{rate}_i= \beta_0 + \beta_1 \times \mathrm{temp}_i + \beta_2 \times 1 [\mathrm{species}_i = \mathrm{O.niveus}] + \epsilon_i$$
where $\epsilon_i \sim N(0,\sigma^2)$ is the error term. 

The fitted model is
$$\mathrm{rate}_i = -7.21 + 3.60 \times \mathrm{temp}_i - 10.07 \times 1[\mathrm{species}_i = \mathrm{O.niveus}]$$

Yes, even though we fitted using lm, this is an **additive ANCOVA model** with **one continuous covariate (temp)**, **one categorical factor (species)**, and **no interaction** between them. **Geometrically**, it fits two parallel regression lines (**one per species**) that **share a common slope** (3.60 chirps/°C) but **have different intercepts, offset by a constant 10.07 chirps**.

# Question 8
Plot the predictions of the model with an additive effect over the data. Is there evidence that the additive effect should be an interaction? Check by fitting the interaction model.

```{r}
crickets |> 
  mutate(pred_additive = predict(ancova_mod)) |>
  ggplot(aes(x = temp, y = rate, color = species)) +
  geom_point() +
  scale_x_continuous("Temperature (°C)") +
  scale_y_continuous("Chirp rate") +
  scale_colour_discrete("Species") +
  geom_line(aes(y = pred_additive)) +
  ggtitle("Additive ANCOVA fit") +
  theme_bw()
```

Now fitting the interaction model
```{r}
interaction_mod <- lm(rate ~ species * temp, data = crickets)
summary(interaction_mod)
```

Compare models using ANOVA()
```{r}
 anova(ancova_mod, interaction_mod)
```

The p value of this test indicates that the interaction is not needed. Formally, **the null hypothesis $H_0: \beta_3 = 0$ where $\beta_3$ is for the interaction effect** of temp:speciesO.niveus, **is not rejected at the $\alpha = 0.05$ level since $0.2542 > 0.05$.**

Visually, the two **parallel** lines from the additive model already track each species’ points closely, and there’s no obvious systematic pattern that would suggest the slopes genuinely differ.

# Question 9
Why might a linear model be inappropriate for the data?

Though the linear ANCOVA fits this particular dataset very well $R^2 \approx 0.99$
**Chirp rate can’t be negative**, but a linear model will predict negative values of chirp rate for temperatures that are low enough.

Careful use of the model (which is a very good predictor) will avoid any problems with this.

# Question 10
Fit each of the above GLMs and interpret the coefficients of the regression.
```{r}
model_wool <- glm(breaks ~ wool, data = warpbreaks, family = poisson(link = "log"))
model_tension <- glm(breaks ~ tension, data = warpbreaks, family = poisson(link = "log"))
model_wool_tension_add <- glm(breaks ~ wool + tension, data = warpbreaks, family = poisson(link = "log"))

summary(model_wool)
summary(model_tension)
summary(model_wool_tension_add)
```
On the log scale, wool As mean is 3.435; wool Bs mean is 0.206 lower. Exponentiating to get back to count scales, 31.0 mean breaks for wool A, 0.814 times fewer breaks for wool B (18.6%). 
The wool and tension coefficients are (as expected for an additive model) essentially unchanged from their single-predictor models. The effects (additive on log scale, multiplicative on raw/count scale) are interpreted holding the other factors fixed: at any given tension, wool B has 18.6% fewer breaks than wool A; at any given wool, Medium tension has 27% less breaks than Low whilst Medium has 40% less.

# Question 11
Write out the fitted model equations.

for model_wool,
$$\mathrm{log(breaks)}_i = 3.435 - 0.206 \times 1[\mathrm{wool}_i = B]$$

for model_tension
$$\mathrm{log(breaks)}_i = 3.594 - 0.321 \times 1[\mathrm{tension}_i = M] - 0.519 \times 1[\mathrm{tension}_i = H]$$

for model_wool_tension_add
$$\mathrm{log(breaks)}_i = 3.692 - 0.206 \times 1 [\mathrm{wool_i} = B] - 0.321 \times 1[\mathrm{tension_i} = M] - 0.519 \times 1 [\mathrm{tension}_i = H]$$

with reference levels $\mathrm{wool}_i = A$ and $\mathrm{tension}_i = L$ in each case. Can be written without the log by exponentiating.